DynaProg for Scala

### 2013-01-17 docs/report/design.tex
Nits.

The support of these features has the following implications:\ul
\item \textbf{Dependency analysis:} Since we target GPUs (and FPGAs) which are massively parallel architecture, a top-down execution using hash tables is impractical (fallback computation if element is not present is hard to parallelize), hence we need to construct the result tree bottom-up, therefore ensure that the (partial) evaluation order between rules is respected.
\item \textbf{Normalization:} in order to automate the backtracking, we need the grammar rules to present a certain shape so that we can define uniquely the backtracking information (in particular we want to distinguish between alternatives). Also we need to maintain coherence between the Scala and the CUDA version so that they can inter-operate: we would like to reuse the backtracking information (from CUDA) to do actual processing in Scala (pretty-printing or actual computation as described in \S\ref{intro}).
\item \textbf{Optimizations:} Since ADP exposes a grammar, we have the possibility to do grammar optimizations (see \S\ref{grammar_opt}). Also since the grammar might define useless rules, we might want to eliminate them: dead rule elimination is very similar to traditional dead code elimination (it reduces the size of generated code) but also reduces the memory consumption (as storage matrix does not need to be reserved) and even speeds up the computation, (since all grammar rules for a particular element are computed at the same time in CUDA implementation, see \ref{dead_rules}).
\ule

@@

Given $f$ , find a pair of functions $(f_1,f_2)$ or $(f_3,f_4)$ such that\footnote{The first equation denotes breaking in two functions, the second is Bellman's optimality principle.}
\[\begin{array}{rcll}
f(i,k_1,k_2,j) &=& f_1(i,f_2(i,k_1,k_2),k_2,j) & \land \\
	\min\limits_{i<k_1<k_2<j}\big[ f(i,k_1,k_2,j) \big] &=& \min\limits_{i<k_2<j} \big[ f_1(i,\min\limits_{i<k_1<k_2} \big[ (f_2(i,k_1,k_2  ) \big],k_2,j) \big] & \lor \vspace{12pt} \\
f(i,k_1,k_2,j) &=& f_3(i,k_1,f_4(k_1,k_2,j),j) & \land \\
	\min\limits_{i<k_1<k_2<j}\big[ f(i,k_1,k_2,j) \big] &=& \min\limits_{i<k_2<j} \big[ f_3(i,k_1,\min\limits_{k_1<k_2<j} \big[ (f_4(k_1,k_2,j) \big],j) \big]
\end{array}\]
% Viktor Kuncak: We can solve this more easily if functions are linear
Since this requires complex mathematical analysis that is out of the scope of the project, and since we have not found relevant literature on that particular subject, we leave this optimization to the responsibility of the user by assuming that the provided grammar is already optimal.

@@

Ad-hoc C translation seems more appropriate than using LMS for the following reasons: \ul
\item LMS uses the type {\tt Rep[T]} to represent a lifted object of type {\tt T}. We tried to use LMS to generate both the Scala and CUDA program. This approach suffered from multiple complication from types, in particular it becomes very tedious to break a lifted tuple into a tuple of lifted values (that is convert {\tt Rep[(A,B,C)]} into {\tt (Rep[A],Rep[B],Rep[C])}), and this approach also limits the functionality of Scala parsers (as only a subset of the Scala language can be lifted into LMS).

@@

From an engineering perspective although the features offered by LMS seem very desirable, we should refrain from adding unnecessary levels of complexity\footnote{KISS engineering principle: \url{http://en.wikipedia.org/wiki/KISS_principle}}. Also given the short time frame allowed to this project (with respect to the requirements), I considered that it was preferable to spend time in making progress in our project rather than helping improving and adding features to an existing project (but I would reconsider my position given different deadlines). Also during our attempts to move all the parsers on LMS, we also investigated in the Scala 2.10 macros (that operate on the Scala abstract syntax tree), which represent a possible alternative to LMS for generating  simple functions for a costing algebra.

>>>>

The support of these features has the following implications:\ul
\item \textbf{Dependency analysis:} Since we target GPUs (and FPGAs) which are massively parallel architecture, a top-down execution using hash tables is impractical (fallback computation if element is not present is hard to parallelize), hence we need to construct the result tree bottom-up, therefore ensure that the (partial) evaluation order between rules is respected.
\item \textbf{Normalization:} in order to automate the backtracking, we need the grammar rules to present a certain shape so that we can define uniquely the backtracking information (in particular we want to distinguish between alternatives). Also we need to maintain coherence between the Scala and the CUDA version so that they can inter-operate: we would like to reuse the backtracking information (from CUDA) to do actual processing in Scala (pretty-printing or actual computation as described in \S\ref{intro}).
\item \textbf{Optimizations:} Since ADP exposes a grammar, we might have the opportunity to do optimizations at the grammar level (see \S\ref{grammar_opt}). Also since the grammar might define useless rules, we might want to eliminate them: dead rule elimination is very similar to traditional dead code elimination (it reduces the size of generated code) but also reduces the memory consumption (as storage matrix does not need to be reserved) and even speeds up the computation, (since all grammar rules for a particular element are computed at the same time in CUDA implementation, see \ref{dead_rules}).
\ule

@@

Given $f$ , find a pair of functions $(f_1,f_2)$ or $(f_3,f_4)$ such that\footnote{The first and third equations denote breaking in two distinct functions, the second and fourth represent Bellman's optimality principle preservation.}
\[\begin{array}{rcll}
f(i,k_1,k_2,j) &=& f_1(i,f_2(i,k_1,k_2),k_2,j) & \land \\
	\min\limits_{i<k_1<k_2<j}\big[ f(i,k_1,k_2,j) \big] &=& \min\limits_{i<k_2<j} \big[ f_1(i,\min\limits_{i<k_1<k_2} \big[ (f_2(i,k_1,k_2  ) \big],k_2,j) \big] & \lor \vspace{12pt} \\
f(i,k_1,k_2,j) &=& f_3(i,k_1,f_4(k_1,k_2,j),j) & \land \\
	\min\limits_{i<k_1<k_2<j}\big[ f(i,k_1,k_2,j) \big] &=& \min\limits_{i<k_2<j} \big[ f_3(i,k_1,\min\limits_{k_1<k_2<j} \big[ (f_4(k_1,k_2,j) \big],j) \big]
\end{array}\]

% Viktor Kuncak: We can solve this more easily if functions are linear
Since this requires complex mathematical analysis that is out of the scope of the project, and since we have not found relevant literature on that particular subject, we leave this optimization to the responsibility of the user by assuming that the provided grammar is already optimal.

Note that this optimization needs to use informations from both the grammar (the rule to split) and the algebra ($f$), which restricting its application to either single algebra or algebra that could share the same optimization, because modifying the grammar will change the associated backtrack information (thereby breaking its compatibility for usage with other algebrae).

@@

Ad-hoc C translation seems more appropriate than using LMS for the following reasons: \ul
\item LMS uses the type {\tt Rep[T]} to represent a lifted object of type {\tt T}. We tried to use LMS to generate both the Scala and CUDA program. This approach suffered from multiple complication from types, in particular it becomes very tedious to break a lifted tuple into a tuple of lifted values (that is convert {\tt Rep[(A,B,C)]} into {\tt (Rep[A],Rep[B],Rep[C])}). This issue is particularly annoying in our particular case because concatenation construct pair of result (hence multiple concatenation would produce nested pairs) whereas we want to present its result as a flattened list of arguments suitable for immediate use by a function.

This approach also complicates the usage of Scala parsers because the algebra can only contain a subset of the Scala language can be represented in LMS. Taking the example of matrix multiplication, the user would like not only to solve the dynamic problem on GPU, but also to leverage the graphic card to compute matrices products. Since associated resources description cannot be represented by LMS nodes, the user would need to extend LMS with ad-hoc DSL for every particular function that is not representable in LMS.

@@

From an engineering perspective although the features offered by LMS seem very desirable, we should refrain from adding unnecessary levels of complexity\footnote{KISS engineering principle: \url{http://en.wikipedia.org/wiki/KISS_principle}}. Also given the short time frame allowed to this project (with respect to the requirements), I considered that it was preferable to spend time in making progress rather than improving an existing project (but I would reconsider my position given different deadlines). Also during our attempts to move all the parsers on LMS, we also investigated in the Scala 2.10 macros\footnote{(Macros operate directly on the internal representation of the Scala compiler, after the type analysis and and before the code generation phase.} \cite{scala_macros}, which represent a possible alternative to LMS for generating simple functions for a costing algebra.

### 2013-01-17 docs/report/introduction.tex
Nits.

% ------------------------------------------------------------------------------------------------
\newpage
\subsection{Scala} \label{bg_scala}
%  - However, we would list some ... => ; this is of course not the only reason; Scala offers some interesting features which make it a nice language to develop a DSL in ?
%  - large adoption in the industry is not really a feature is it?
%  - Implicit functions definition => implicit functions and parameters offer a way to neatly integrate primitive/pre-existing constructs into a DSL, by converting their types in order to correspond to the DSL semantics, while at the same time guarantee-ing type safety.
%  - Macros[6] and LMS are mechanisms to modify semantics of a program; in particular, these tools can be used to implement domain-specific optimizations of programs. LMS also contains a code generation platform which we use to generate some functions in C (link to section).
%  - Mixin composition (traits) allow for multiple inheritance and composition, which makes it easy to represent, in our DSL, multiple algebrae, and also interface with different runtimes (link to section).
\textit{<<Scala is a general purpose programming language designed to express common programming patterns in a concise, elegant, and type-safe way. It smoothly integrates features of object-oriented and functional languages, enabling programmers to be more productive. Many companies depending on Java for business critical applications are turning to Scala to boost their development productivity, applications scalability and overall reliability.>>}\footnote{\url{http://www.scala-lang.org/node/25}}

As the Scala \cite{scala} programming language is developed by our laboratory (LAMP, EPFL), it seems natural to use it as host language for our project, however, we would list some of its features \cite{scala_api} that makes it an interesting development language for this project:\ul
\item The functional programming style and syntactic sugar offered by Scala allow concise writing of implementation, analysis and transformations of our DSL, allowing us to focus on \textit{what} we want to achieve instead of \textit{how}.
\item Scala is largely adopted in the industry\footnote{\url{http://www.scala-lang.org/node/1658}}, which makes both the adoption of related project easier and reduces the learning time of its potential users.
\item Since Scala programs execute in the Java Virtual Machine (JVM), they can benefit of the native interface (JNI) that offers the possibility to dynamically load libraries (usually written in C) and possibly interact with CUDA to leverage the GPU.
\item Scala is equipped with a strong typing and type inference system that reduces the syntactical constraints while putting strong guarantees on type correctness at compilation.
\item Implicit functions definition can help interfacing functions by automatically changing the type of passed arguments
\item Manifests (or TypeTags and ClassTags) allow type extraction at runtime (we use this to convert a Scala type into a C/CUDA type)
\item Macros\cite{scala_macros} and LMS (\S\ref{bg_lms}) could be used to modify the semantics of a specific part of the user program (we currently use LMS to produce the corresponding C function).
\item One Scala concept that we heavily use is \textit{traits} that can be viewed as abstract classes and mixed together, thereby allowing multiple inheritance.
\ule

@@

A discussion of the integration of LMS in our project can be found in \S\ref{lms_use}.

>>>>

% ------------------------------------------------------------------------------------------------
\newpage
\subsection{Scala} \label{bg_scala}
\textit{<<Scala is a general purpose programming language designed to express common programming patterns in a concise, elegant, and type-safe way. It smoothly integrates features of object-oriented and functional languages, enabling programmers to be more productive. Many companies depending on Java for business critical applications are turning to Scala to boost their development productivity, applications scalability and overall reliability.>>}\footnote{\url{http://www.scala-lang.org/node/25}}

As the Scala \cite{scala} programming language is developed by our laboratory (LAMP, EPFL), it seems natural host language for our project. Its large adoption\footnote{\url{http://www.scala-lang.org/node/1658}}, would make the adoption of our DSL easier while reducing the learning time of its potential users. Additionally, some features \cite{scala_api} of Scala makes it an interesting development language for this project:\ul
\item The functional programming style and syntactic sugar offered by Scala allow concise writing of implementation, analysis and transformations of our DSL, allowing us to focus on \textit{what} we want to achieve instead of \textit{how}.
\item Since Scala programs execute in the Java Virtual Machine (JVM), they can benefit of the native interface (JNI) that offers the possibility to dynamically load libraries (usually written in C) and possibly interact with CUDA to leverage the GPU.
\item Scala is equipped with a strong typing and type inference system that reduces the syntactical constraints while putting strong guarantees on type correctness at compilation.
\item Implicit functions and parameters allow to simplify the syntactic usage of the DSL by implementing automatic conversions, while at the same time preserving type safety.
\item Manifests (or TypeTags and ClassTags) allow type extraction at runtime (we use this to convert a Scala type into a C/CUDA type)
\item Macros\cite{scala_macros} and LMS (\S\ref{bg_lms}) could be used to modify the semantics of specific parts, or implement domain-specific optimizations of the user program. LMS also contains a multi-language code generator that we leverage to produce C functions (see \S\ref{user_fun_gen}).
\item One Scala concept that we heavily use is \textit{traits} that can be viewed as abstract classes and combined (mixin composition), thereby allowing multiple inheritance. We use this feature in particular to smoothly combine algebra, grammar and possibly code generation (\S\ref{codegen}) into a concrete program.
\ule

@@

A discussion on the integration of LMS in our project can be found in \S\ref{lms_use}.

### 2013-01-17 docs/report/problems.tex
Nits.

\newpage
\section{Dynamic programming problems} \label{problems}
There exist various categories of dynamic programming:\ul
\item Series that operate usually on a single dimension (like Fibonacci\footnote{http://en.wikipedia.org/wiki/Fibonacci_number})
\item Sequences alignment (matching two sequences at best), top-down grammar analysis (parenthesizing), sequence folding, ... (see \S\ref{problems} for more examples and detailed classification)
\item Tree-related algorithms: phylogeny \cite{philogeny}, trees raking \cite{tree_raking}, maximum tree independent set \cite{tree_independent_set}, ... (can be viewed as a sparse version of the second category)
\ule

@@

% ----------------------------------------------
\subsubsection{Definitions}\ul
\item \textbf{Cost or score:} refers to the result of the dynamic programming recurrence formula.
\item \textbf{Backtrack:} the backtrack is the information related to a score that describe how it has been obtained by referring to immediately previous elements. By induction on the backtrack, the \textbf{trace} (that describe \textit{all} thee steps to obtain the result) can be obtained.
\item \textbf{Alphabets:} an alphabet is an enumeration of the possible values. Its size helps determining how many bits are required in the implementation to store all values. Alphabets are defined for input, cost, backtrack and wavefront.
\item \textbf{Dimensions:} let $n$ the size of the input and $d$ the dimension of the underlying matrix.
\item \textbf{Matrices:} we refer by \textit{matrix} or \textit{matrices} to all the memoized intermediate cost- and backtrack-related information that is necessary to solve the dynamic programming problem of interest. Matrix elements are usually denoted by $M_{(i,j)}$ ($i^{\rm th}$ line , $j^{\rm th}$ column).
\item \textbf{Computation block:} this is a part of the DP matrix (cost and backtrack) that we want to compute. A block might be either a sub-matrix (rectangular) or a parallelogram (possibly reduced by taking the intersection with its enclosing matrix).
\item \textbf{Wavefront:} the wavefront consists of the minimum data necessary to construct a computation block of the DP matrix. It might include some previous lines/columns/diagonals as well as line-/column-/diagonal-wise aggregations (min, max, sum, ...).
\item \textbf{Delay:} we call delay the maximum distance between an element and its dependencies along column and lines (ex: recurrence $M_{(i,j)}=f\big(M_{(i+1,j)}, M_{(i+2,j-1)}\big)$ has delay 3).
\ule

@@

% ------------------------------------------------------------------------------------------------
\subsection{Problems of interest} \label{problem_examples}
We here focus on problems that have an underlying bi-dimensional matrix ($d=2$) because they can be parallelized (as opposed to be serial if $d=1$) and can solve large problems (of size $n$). Problems of higher matrix dimensionality ($d\ge3$) require substantial memory which severely impacts their scalability. Also it seems that most problems of interest have an algorithmic complexity of at most $O(n^4)$, probably because running time would otherwise becomes a severely limiting factor for the size of the problem.

@@

% ----------------------------------------------
\newpage
\subsubsection{Nussinov algorithm} \label{nussinov} \ol
\item Problem: folding a RNA string $S$ over itself $|S|=n$.
\item Matrices: $M_{n\times n}, B_{n \times n}$
\item Alphabets:\ul
	\item Input: $\Sigma(S)=\{a,c,g,u\}$.
	\item Cost matrix: $\Sigma(M)=\{0..n\}$
	\item Backtrack matrix: $\Sigma(B)=\{stop,D,1..n\}$
	\ule
\item Initialization: \ul
	\item Cost matrix: $ M_{(i,i)}=M_{(i,i+1)}=0$
	\item Backtrack matrix: $B_{(i,i)}=B_{(i,i+1)}=stop$
	\ule
\item Recurrences:
	\[M_{(i,j)}=\max\left\{\begin{array}{l|l}
		M_{(i+1,j-1)}+\omega(i,j) & D\\
		\max_{i\le k<j}M_{(i,k)}+M_{(k+1,j)} & k
	\end{array}\right\} = B_{(i,j)} \]
	With $\omega(i,j)=1$ if $i,j$ are complementary. 0 otherwise.
\item Backtracking: Start the backtracking in $B_{(0,n-1)}$ and go backward. The backtracking is very similar to that of the matrix multiplication, except that we also introduce the diagonal matching.
\item Visualization:
\CFig{Nussinov dependencies}{
	\put(-.7,6.5){\rotatebox{90}{First}}\put(-.4,6.4){\linethickness{1pt}\vector(0,-1){2}}
	\put(.2,8.2){Last}\put(1.5,8.4){\linethickness{1pt}\vector(1,0){2}}
	\Cfl{\Cg}\Cfd{\Cz}\Cfdu{\Cz}
	\Cd[0,1]{6,1}{2.8}\Cd[0,1]{6,2}{1.8}\Cd[0,1]{6,3}{0.8}
	\Cd[1,0]{3,4}{2.8}\Cd[1,0]{4,4}{1.8}\Cd[1,0]{5,4}{0.8}
	\Cd[1,1]{5,3}{0.8}
	\put(3.5,4.5){\line(3,-1){3}}\put(4.5,4.5){\line(2,-2){2}}\put(5.5,4.5){\line(1,-3){1}}
}

@@

% ----------------------------------------------
\newpage
\subsubsection{Zuker RNA folding} \label{zuker} \ol
\item Problem: folding a RNA string $S$ over itself $|S|=n$ by minimizing the free energy.
\item Matrices: $V_{n\times n}, W_{n\times n}, F_n$ (Free Energy),  $BV_{n \times n}, BW_{n \times n}, BF_n$
\item Alphabets:\ul
	\item Input: $\Sigma(S)=\{a,c,g,u\}$.
	\item Cost matrices:\ul
		\item $\Sigma(W)=\Sigma(V)=\{0..z\}$ with $z \le n \cdot b+c$
		\item $\Sigma(F)=\{0..y\}$ with $y\le \min(F_0, z\cdot n)$
		\ule
	\item Backtrack matrices: \ul
		\item $\Sigma(BW)=\{stop, L,R,V,k\}$
		\item $\Sigma(BV)=\{stop, hairpin, stack, (i,j) , k\}$ with $0\le i,j,k < n$ % \\ $HL$=HairpinLoop, $IL$=InteriorLoop, $(i,j)$=MultiLoop
		\item $\Sigma(BF)=\{stop, P, k\}$ with $0\le k < n$
		\ule
	\ule
\item Initialization:\ul
	\item Cost matrices: $W_{(i,i)}=V_{(i,i)}=0, F_{(0)}=$ energy of the unfolded RNA.
	\item Backtrack matrices: $BW_{(i,i)}=BV_{(i,i)}=BF_{(0)}=stop$.
	\ule
\item Recurrence:
\[\begin{array}{rcl}
W_{(i,j)}&=&\min\left\{\begin{array}{l|l}
	W_{(i+1,j)}+b & L\\
	W_{(i,j-1)}+b & R\\
	V_{(i,j)}+\delta(S_i,S_j) & V \\
	\min_{i<k<j}W_{(i,k)}+W_{(k+1,j)} &k
\end{array}\right\} = BW_{(i,j)}\\
\\
V_{(i,j)}&=&\min\left\{\begin{array}{l|l}
	\infty \qquad\qquad\qquad\qquad {\rm if}(S_i,S_j) \text{ is not a base pair} & stop\\\\
	eh(i,j)+b \qquad\qquad\, \text{otherwise} & hairpin\\
	V_{(i+1,j-1)}+es(i,j) & stack \\
	VBI_{(i,j)} & (i',j') \\
	\min_{i<k<j-1}\{W_{(i+1,k)}+W_{(k+1,j-1)}\} +c & k
\end{array}\right\} = BV_{(i,j)}\\
\\
VBI_{(i,j)}&=&\min\Big\{\min_{i<i'<j'<j}V_{(i',j')}+ebi(i,j,i',j')\} +c \,\,\Big|\,\, (i',j') \Big\}=BV_{(i,j)}\\
\\
F_{(j)}&=&\min\left\{\begin{array}{l|l}
	F_{(j-1)} & P \\
	\min_{1\le i< j} (F_{(i-1)} + V_{(i,j)}) & i
\end{array}\right\} = BF_{(j)}
\end{array}\]

@@

\item Backtracking: Start at $BF_{(n)}$ using the recurrences
 \[\begin{array}{rcl}
	BF_{(j)} &=& \left\{\begin{array}{rcl} P&\implies& BF_{(j-1)} \\ i &\implies& BF_{(i-1)} + BV_{(i,j)} \end{array} \right.\\
	\\
	BV_{(i,j)} &=& \left\{\begin{array}{rcl}
		hairpin &\implies&\big< {\rm hairpin}(i,j) \big> \\
		stack &\implies& \big< {\rm stack}(i,j) \big> \oplus BV_{(i+1,j-1)} \\
		(i',j') &\implies& \big< \text{bulge from $(i,j)$ to }(i',j') \big> \oplus BV(i',j')\\
		k &\implies& BW_{(i+1,k)} \oplus BW_{(k+1,j-1)}
	\end{array}\right.\\
	\\
	BW_{(i,j)} &=& \left\{\begin{array}{rcl}
	L & \implies & \big< multi\_loop(i) \big> \oplus BW_{(i+1,j)} \\
	R & \implies & \big< multi\_loop(j) \big> \oplus BW_{(i,j+1)} \\
	V &\implies& BV_{(i,j)} \\
	k &\implies& BW_{(i+1,k)} \oplus BW_{(k+1,j-1)}
	\end{array}\right.
\end{array}\]

@@

The recurrence consists of two non-serial dependencies as in \S\ref{swat_arbitrary} plus a bounded 2-dimensional dependency for bulges.\\[12pt]
Since this problem is non-trivial to understand from the recurrences, we propose an additional illustration of a RNA chain folded according to the Zuker folding algorithm.
\begin{figure}[H]\begin{center}
\includegraphics[width=10cm]{inc/zuker_struct.pdf} \\
\small Types of structural features modeled by the Zuker folding algorithm include: dangling ends (1), \\ internal loop (11), stack (23), multi-loop (47), bulge (68) and hairpin loop (78).\end{center}\caption{An example of an RNA folded into a secondary structure}\end{figure}

>>>>

\newpage
\section{Dynamic programming problems} \label{problems}
There exist various categories of dynamic programming:\ul
\item Series that operate usually sequentially on a single dimension (like Fibonacci\footnote{\url{http://en.wikipedia.org/wiki/Fibonacci\_number}})
\item Sequences alignment (matching two sequences at best), top-down grammar analysis (parenthesizing), sequence folding, ... (see \S\ref{problems} for more examples and detailed classification)
\item Tree-related algorithms: phylogeny \cite{philogeny}, trees raking \cite{tree_raking}, maximum tree independent set \cite{tree_independent_set}, ... (can be viewed as a sparse version of the second category)
\ule

@@

% ----------------------------------------------
\subsubsection{Definitions}\ul
\item \textbf{Cost or score:} refers to the result of the dynamic programming recurrence formula.
\item \textbf{Backtrack:} the backtrack is the information related to a score that describe how it has been obtained by referring to immediately previous elements. By induction on the backtrack, the \textbf{trace} (that describe \textit{all} thee steps to obtain the result) can be obtained.
\item \textbf{Alphabets:} an alphabet is an set of possible values. Its size helps determining how many bits are required in the implementation to represent all its elements. Alphabets are defined for input, cost, backtrack and wavefront.
\item \textbf{Dimensions:} let $n$ the size of the input and $d$ the dimension of the underlying matrix.
\item \textbf{Matrices:} we refer by \textit{matrix} or \textit{matrices} to all the memoized intermediate cost- and backtrack-related information that is necessary to solve the dynamic programming problem of interest. Matrix elements are usually denoted by $M_{(i,j)}$ ($i^{\rm th}$ line , $j^{\rm th}$ column).
\item \textbf{Computation block:} this is a part of the DP matrix (cost and backtrack) that we want to compute. A block might be either a sub-matrix (rectangular) or a parallelogram (possibly reduced by taking the intersection with its enclosing matrix).
\item \textbf{Wavefront:} the wavefront consists of the minimum data necessary to construct a computation block of the DP matrix. It might include some previous lines/columns/diagonals as well as line-/column-/diagonal-wise aggregations (min, max, sum, ...).
\item \textbf{Delay:} we call delay the maximum distance between an element and its dependencies along column and lines (ex: recurrence $M_{(i,j)}=f\big(M_{(i+1,j)}, M_{(i+2,j-1)}\big)$ has delay 3).
\ule

@@

% ------------------------------------------------------------------------------------------------
\newpage
\subsection{Problems of interest} \label{problem_examples}
We here focus on problems that have an underlying bi-dimensional matrix ($d=2$) because they can be parallelized (as opposed to be serial if $d=1$) and can solve large problems (of size $n$). Problems of higher matrix dimensionality ($d\ge3$) require substantial memory which severely impacts their scalability. Also it seems that most problems of interest have an algorithmic complexity of at most $O(n^4)$, probably because running time would otherwise becomes a severely limiting factor for the size of the problem.

@@

% ----------------------------------------------
\newpage
\subsubsection{Nussinov algorithm} \label{nussinov} \ol
\item Problem: folding a RNA string $S$ over itself $|S|=n$, according to matching properties ($\omega$) of its elements (also called bases).
\item Matrices: $M_{n\times n}, B_{n \times n}$
\item Alphabets:\ul
	\item Input: $\Sigma(S)=\{a,c,g,u\}$.
	\item Cost matrix: $\Sigma(M)=\{0..n\}$
	\item Backtrack matrix: $\Sigma(B)=\{stop,D,1..n\}$
	\ule
\item Initialization: \ul
	\item Cost matrix: $ M_{(i,i)}=M_{(i,i+1)}=0$
	\item Backtrack matrix: $B_{(i,i)}=B_{(i,i+1)}=stop$
	\ule
\item Recurrences:
	\[M_{(i,j)}=\max\left\{\begin{array}{l|l}
		M_{(i+1,j-1)}+\omega(i,j) & D\\
		\max_{i\le k<j}M_{(i,k)}+M_{(k+1,j)} & k
	\end{array}\right\} = B_{(i,j)} \]
	With $\omega(i,j)=1$ if $i,j$ are complementary. 0 otherwise.
\item Backtracking: Start the backtracking in $B_{(0,n-1)}$ and go backward. The backtracking is very similar to that of the matrix multiplication, except that we also introduce the diagonal matching.
\item Visualization:
\CFig{Nussinov dependencies}{
	\put(-.7,6.5){\rotatebox{90}{First}}\put(-.4,6.4){\linethickness{1pt}\vector(0,-1){2}}
	\put(.2,8.2){Last}\put(1.5,8.4){\linethickness{1pt}\vector(1,0){2}}
	\Cfl{\Cg}\Cfd{\Cz}\Cfdu{\Cz}
	\Cd[0,1]{6,1}{2.8}\Cd[0,1]{6,2}{1.8}\Cd[0,1]{6,3}{0.8}
	\Cd[1,0]{3,4}{2.8}\Cd[1,0]{4,4}{1.8}\Cd[1,0]{5,4}{0.8}
	\Cd[1,1]{5,3}{0.8}
	\put(3.5,4.5){\line(3,-1){3}}\put(4.5,4.5){\line(2,-2){2}}\put(5.5,4.5){\line(1,-3){1}}
}

@@

% ----------------------------------------------
\newpage
\subsubsection{Zuker RNA folding} \label{zuker} \ol
\item Problem: folding a RNA string $S$ over itself $|S|=n$ by minimizing the free energy (which is based on actual measurements, hereby much more complicated than Nussinov \S\ref{nussinov}).
\item Matrices: $V_{n\times n}, W_{n\times n}, F_n$ (Free Energy),  $BV_{n \times n}, BW_{n \times n}, BF_n$
\item Alphabets:\ul
	\item Input: $\Sigma(S)=\{a,c,g,u\}$.
	\item Cost matrices:\ul
		\item $\Sigma(W)=\Sigma(V)=\{0..z\}$ with $z \le n \cdot b+c$
		\item $\Sigma(F)=\{0..y\}$ with $y\le \min(F_0, z\cdot n)$
		\ule
	\item Backtrack matrices: \ul
		\item $\Sigma(BW)=\{stop, L,R,V,k\}$
		\item $\Sigma(BV)=\{stop, hairpin, stack, (i,j) , k\}$ with $0\le i,j,k < n$ % \\ $HL$=HairpinLoop, $IL$=InteriorLoop, $(i,j)$=MultiLoop
		\item $\Sigma(BF)=\{stop, P, k\}$ with $0\le k < n$
		\ule
	\ule
\item Initialization:\ul
	\item Cost matrices: $W_{(i,i)}=V_{(i,i)}=0, F_{(0)}=$ energy of the unfolded RNA.
	\item Backtrack matrices: $BW_{(i,i)}=BV_{(i,i)}=BF_{(0)}=stop$.
	\ule
\item Recurrence:
\[\begin{array}{rcl}
W_{(i,j)}&=&\min\left\{\begin{array}{l|l}
	W_{(i+1,j)}+b & L\\
	W_{(i,j-1)}+b & R\\
	V_{(i,j)}+\delta(S_i,S_j) & V \\
	\min_{i<k<j}W_{(i,k)}+W_{(k+1,j)} &k
\end{array}\right\} = BW_{(i,j)}\\
\\
V_{(i,j)}&=&\min\left\{\begin{array}{l|l}
	\infty \qquad\qquad\qquad\qquad {\rm if}(S_i,S_j) \text{ is not a base pair} & stop\\\\
	eh(i,j)+b \qquad\qquad\, \text{otherwise} & hairpin\\
	V_{(i+1,j-1)}+es(i,j) & stack \\
	VBI_{(i,j)} & (i',j') \\
	\min_{i<k<j-1}\{W_{(i+1,k)}+W_{(k+1,j-1)}\} +c & k
\end{array}\right\} = BV_{(i,j)}\\
\\
VBI_{(i,j)}&=&\min\Big\{\min_{i<i'<j'<j}V_{(i',j')}+ebi(i,j,i',j')\} +c \,\,\Big|\,\, (i',j') \Big\}=BV_{(i,j)}\\
\\
F_{(j)}&=&\min\left\{\begin{array}{l|l}
	F_{(j-1)} & P \\
	\min_{1\le i< j} (F_{(i-1)} + V_{(i,j)}) & i
\end{array}\right\} = BF_{(j)}
\end{array}\]

@@

\item Backtracking: starts at $BF_{(n)}$ and uses the recurrences
 \[\begin{array}{rcl}
	BF_{(j)} &=& \left\{\begin{array}{rcl} P&\implies& BF_{(j-1)} \\ i &\implies& BF_{(i-1)} + BV_{(i,j)} \end{array} \right.\\
	\\
	BV_{(i,j)} &=& \left\{\begin{array}{rcl}
		hairpin &\implies&\big< {\rm hairpin}(i,j) \big> \\
		stack &\implies& \big< {\rm stack}(i,j) \big> \oplus BV_{(i+1,j-1)} \\
		(i',j') &\implies& \big< \text{bulge from $(i,j)$ to }(i',j') \big> \oplus BV(i',j')\\
		k &\implies& BW_{(i+1,k)} \oplus BW_{(k+1,j-1)}
	\end{array}\right.\\
	\\
	BW_{(i,j)} &=& \left\{\begin{array}{rcl}
	L & \implies & \big< multi\_loop(i) \big> \oplus BW_{(i+1,j)} \\
	R & \implies & \big< multi\_loop(j) \big> \oplus BW_{(i,j+1)} \\
	V &\implies& BV_{(i,j)} \\
	k &\implies& BW_{(i+1,k)} \oplus BW_{(k+1,j-1)}
	\end{array}\right.
\end{array}\]

@@

The recurrence consists of two non-serial dependencies as in \S\ref{swat_arbitrary} plus a bounded 2-dimensional dependency for bulges.

\newpage
Since this problem is non-trivial to understand from the recurrences, we propose an additional illustration of a RNA chain folded according to the Zuker folding algorithm.
\begin{figure}[H]\begin{center}
\includegraphics[width=10cm]{inc/zuker_struct.pdf} \\
\small Types of structural features modeled by the Zuker folding algorithm include: dangling ends (1), \\ internal loop (11), stack (23), multi-loop (47), bulge (68) and hairpin loop (78).\end{center}\caption{An example of an RNA folded into a secondary structure}\end{figure}
